Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-03-08 paper.tex
oka

Correct identification of protein posttranslational-modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\cite{Li_2016} is a recently developed tool for global identification of PTMs.
Spectral file calibration prior to applying the tool, and algorithmic improvements in the peptide database search greatly improve the accuracy and efficiency of PTM identification.
We enhanced G-PTM-D by using targeted mass-window searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The proposed changes lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Additional advantages of using targeted mass-windows searches in context of novel modification discovery are shown.
\end{abstract}

@@

A recently described bioinformatics tool for the discovery and localization of new PTMs from "bottom-up" tandem mass spectrometric datasets, G-PTM-D\cite{Li_2016} is the foundation for the work described in this paper.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015,Na_2011} that provides spectral matches to unmodified peptides along with the mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM).
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database.
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

Before describing the proposed changes to G-PTM-D, a few definitions of the various search modes must be introduced.
A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da).
A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account.
We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches.
The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses.
The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches.
The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum).
Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

@@

The mass-window search is useful in other procedures as well, beyond the G-PTM-D workflow.
Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search.
A combination of specialized notch and interval searches produces superior identification and discovery results.

Along with algorithmic improvements, spectral calibration prior to running the G-PTM-D procedure is essential.
A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration, REF).
We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The increased mass accuracy of the recalibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

\begin{figure}
\end{figure}

\begin{figure}
\begin{subfigure}[b]{.5\linewidth}
  \includegraphics{figureDifferentSearchModes.eps}
  \caption{Search Modes}
  \label{fgr:differentSearchModes}
\end{subfigure}%
\begin{subfigure}[b]{.5\linewidth}
  \includegraphics[scale=0.5]{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{subfigure}
\caption{A figure}\label{fig:1}
\end{figure}

@@

We first contrast mass-window searches with narrow-mass and open-mass searches, and demonstrate the advantages of using mass-window searches.
The we demonstrate the improvements in the performance of the enhanced G-PTM-D over the original proposed methods, that are due to using mass-window searches.
Finally, we discuss the application of mass-window searches to the discovery of novel modifications.

@@

@@

We describe the heuristic approach for augmenting the modification database and the final notch list.
After the conclusion of the search, the PSMs within 1\% FDR are selected, and the corresponding mass differences are grouped into peaks, which are categorized into one of the following categories:
\begin{itemize}
\item \textit{No mass error} Exact match (within a certain noise tolerance) between the peptide corresponding to the MS2 fragmentation and the reported precursor monoisotopic mass.
\item \textit{Monoisotopic error} Peaks around values such as 1.003, 2.006, 3.009 (average differences in mass between the monoisotopic peaks and the first few isotope peks). These correspond to having misidentified the monoisotopic peak in the preprocessing deconvolution step. In this case, the identified peptide is still correct. 
\item \textit{Exact PTM or Adduct Masses}e.g. 42.010 for Acetylation or 21.981 for Sodium adduct. In this analysis, there is no conceptual difference between PTMs and adducts: both are observed simply as mass differences between the identified peptide and the precursor mass.  These peaks are further classified into
\begin{itemize}
\item Localizeable - e.g. Methylation. There is evidence for the modification in the MS2 spectrum: the corresponding peaks are shifted by an appropriate amount. 
\item Labile - e.g. Sulfonation. There is no evidence of the modification in the fragmentation spectrum. 
\item Either Localizeable or Labile, e.g. Phosphorylation. Sometimes there is evidence of it in the fragmentation spectrum, and sometimes there is not. 
\end{itemize}
\item Amino acid removals, additions or substitutions.
\item Modification dependent mass shifts - these peaks occur only in presence of certain modifications in the identified peptide. 	E.g. -15.995 in presence of an identified oxidation, or -79.966 in presence on an identified phosphorylation. 
\item Combinations of any of the above, e.g. 1.987 for combination of a monoisotopic error and a deamidation. Some of these combinations occur frequently, and it is crucial to account for them. 
\end{itemize}

An automated script analyzes each identified peak, and provides clues to the nature of the peak. For every peak, a profile is automatically generated.
It includes the total number of unique peptides associated with the mass shift, the fraction of decoys, mass match with any known entry in the Unimod or UniProt database, mass match to an amino acid addition/removal combination, mass match to a combination of higher frequency peaks, fraction of localizable targets, localization residues and/or termini, and presence of any modifications in the matched peptides.
We follow the following procedure to classify mass shift peaks. 
1.	A z score is computed for the fraction of decoys in the peak, comparing with 1\%FDR. If it corresponds to p>0.05, we do not include this peak in the analysis, because there is no sufficient confidence it produces legitimate results.
2.

An attempted localization of the mass difference is automatically performed on every peptide-spectrum match. We carefully examine the mass differences with a localization fraction greater than 0.2, and attempt to deduce the chemical formula, specificity sites and positions within a peptide.
Once a determination has been made, we add the new modification type to the modifications list.
An example output of this step is:

The improvements of using a Comb search instead of an Open search are two-fold. We summarize the differences in Table 2.
	Open Search	Comb Search
Search Time	54.36 hrs	29.66 hrs

The improvement in the discernibility of PTMs is apparent when running a comb search on calibrated vs uncalibrated spectra files. Specifically, the discernibility of PTMs with similar mass errors (ones that are only different because of the mass defect).
This was not really possible previously.
In the current run, Sulfation and Phosphorylation are readily distinguishable.
We present the numbers for the example below in the step 5 numerical validation section.
OPEN SEARCH CASE STUDY
Introduction of the comb search instead of an open search is motivated by a careful review of the Unimod database.
Known modifications with mass difference within 200 Daltons have values that are within [-0.1, 0.2] of every integer.
This interval choice stems from the mass defect in the primary isotope of the common elements.
PTM combinations also have this property.
We tested multiple search strategies in a search for one that gives the smallest number of false positives. 
\subsubsection{Known Modification Confirmation Case Study}

Confirmation of the properties of Sodium Adducts, Phosphorylation and Sulfonation.
\subsubsection{Glycans Case Study}
Proteins that contain modifications such as Hexose, HexNAC, and similar ones are interesting.
The uniprot database only lists some possible sites for such modifications, but does not specify the actual type, thus creating ambiguity regarding the type of the modification.
The new modification discovery workflow is well suited to identify such proteins.
Table~\ref{tbl:glyco} lists the automatically identified glyco modification masses and identities, sorted by frequency of occurence.

\begin{table}[]
\centering
\caption{Glyco}
\label{tbl:glyco}
\begin{tabular}{lll}
MassShift & Count & UnimodID                        \\
\hline
203.079   & 30    & HexNAc                          \\
162.052   & 18    & Hex                             \\
876.322   & 10    & Hex(2)HexNAc(2)dHex(1)          \\
656.228   & 8     & Hex(1)HexNAc(1)NeuAc(1)         \\
365.132   & 8     & Hex(1)HexNAc(1)                 \\
1216.422  & 8     & Hex(5)HexNAc(2)                 \\
730.265   & 6     & Hex(2)HexNAc(2)                 \\
1378.475  & 4     & Hex(6)HexNAc(2)                 \\
947.323   & 4     & Hex(1)HexNAc(1)NeuAc(2)         \\
963.317   & 3     & Hex(1)HexNAc(1)NeuAc(1)NeuGc(1) \\
178.047   & 3     & Galactosyl                      \\
1038.375  & 3     & dHex(1)Hex(3)HexNAc(2)          \\
146.058   & 2     & dHex                            \\
340.1     & 2     & Glucosylgalactosyl              \\
406.158   & 2     & HexNAc(2)                       \\
697.256   & 2     & HexNAc(2)NeuAc(1)               \\
672.223   & 2     & Hex(1)HexNAc(1)NeuGc(1)         \\
860.326   & 2     & Hex(1)HexNAc(2)dHex(2)         
\end{tabular}
\end{table}

\subsubsection{Novel Modification Case Study}

%
%\subsection{References}
%
%The class makes various changes to the way that references are
%handled.  The class loads \textsf{natbib}, and also the
%appropriate bibliography style.  References can be made using
%the normal method; the citation should be placed before any
%punctuation, as the class will move it if using a superscript
%citation style
%\cite{Mena2000,Abernethy2003,Friedman-Hill2003,EuropeanCommission2008}.
%The use of \textsf{natbib} allows the use of the various citation
%commands of that package: \citeauthor{Abernethy2003} have shown
%something, in \citeyear{Cotton1999}, or as given by
%Ref.~\citenum{Mena2000}.  Long lists of authors will be
%automatically truncated in most article formats, but not in
%supplementary information or reviews \cite{Pople2003}. If you
%encounter problems with the citation macros, please check that
%your copy of \textsf{natbib} is up to date. The demonstration
%database file \texttt{achemso-demo.bib} shows how to complete
%entries correctly. Notice that ``\latin{et al.}'' is auto-formatted
%using the \texttt{\textbackslash latin} command.
%
%Multiple citations to be combined into a list can be given as
%a single citation.  This uses the \textsf{mciteplus} package
%\cite{Johnson1972,*Arduengo1992,*Eisenstein2005,*Arduengo1994}.
%Citations other than the first of the list should be indicated
%with a star. If the \textsf{mciteplus} package is not installed,
%the standard bibliography tools will still work but starred
%references will be ignored. Individual references can be referred
%to using \texttt{\textbackslash mciteSubRef}:
%``ref.~\mciteSubRef{Eisenstein2005}''.
%
%The class also handles notes to be added to the bibliography.  These
%should be given in place in the document \bibnote{This is a note.
%The text will be moved the the references section.  The title of the
%section will change to ``Notes and References''.}.  As with
%citations, the text should be placed before punctuation.  A note is
%also generated if a citation has an optional note.  This assumes that
%the whole work has already been cited: odd numbering will result if
%this is not the case \cite[p.~1]{Cotton1999}.

%\newpage
%
%\subsection{Floats}
%
%The example file also loads the optional \textsf{mhchem} package, so
%that formulas are easy to input: \texttt{\textbackslash ce\{H2SO4\}}
%gives \ce{H2SO4}.  See the use in the bibliography file (when using
%titles in the references section).
%
\begin{acknowledgement}

@@

A listing of the contents of each file supplied as Supporting Information
should be included.
For instructions on what should be included in the
Supporting Information as well as how to prepare this material for
publications, refer to the journal's Instructions for Authors.

>>>>

Correct identification of protein posttranslational-modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\cite{Li_2016} is a recently developed tool for global identification and localization of PTMs.
Spectral file calibration prior to applying the tool, and algorithmic improvements in the peptide database search greatly improve the accuracy and efficiency of PTM identification.
We enhanced G-PTM-D by using targeted mass-window searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The proposed changes lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Targeted mass-windows searches are also shown to be useful as a replacement to a standard wide-mass search, resulting in a 10\% increase in the number of confidently identified peptides: this leads to improvements in novel modification discovery heuristics originally based on the wide-mass search.
\end{abstract}

@@

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist, e.g. MODa\cite{Na_2011}.
G-PTM-D\cite{Li_2016}, a recently described bioinformatics tool for the discovery and localization of new PTMs, is the foundation for the work described in this paper.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015, Na_2011} that provides spectral matches to unmodified peptides along with the mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM).
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database.
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

As described, G-PTM-D has a few significant limitations.
The initial wide-mass search procedure is slow, taking hours for modest size datasets.
Due to this, high mass modifications, such as glycosolations are out of reach, since widening the search window from the suggested $\pm 200$ Daltons increases the search time even more.
Furthermore, in G-PTM-D, modification discovery is limited to PTMs in the Uniprot database. 
Finally, PTMs that are similar in mass, such as Phosphorylation and Sulfonation are problematic since they are virtually indistinguishable in unprocessed spectra files.

Spectral calibration prior to running the G-PTM-D procedure is essential for distinguishing modifications with similar mass.
As a side effect, calibration signinficantly increases the overall number of confidently identified peptides and modifications.

@@

The idea of restricting the search space by only allowing certain mass differences between the precursor and the identified peptide mass is useful in other procedures as well, beyond the G-PTM-D workflow.
Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search.
A combination of specialized notch and interval searches produces superior identification and discovery results.

\begin{figure}
  \includegraphics[scale=0.5]{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

@@

Post-acquisition calibration is a useful tool that is shown to increase the number of confidently identified peptides, and enables qualitatively distinguishing modifications with similar mass.
Then, mass-window searches are contrasted with narrow-mass and open-mass searches.
The improvements in the performance of the enhanced G-PTM-D over the originally proposed G-PTM-D, that are due to using mass-window searches are shown.
Finally, the application of mass-window searches to the discovery of novel modifications is discussed.

A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration, REF).
We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The increased mass accuracy of the recalibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

@@

A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da).
A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account.
We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches.
The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses.
The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches.
The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum).
Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

\begin{figure}
  \includegraphics{figureDifferentSearchModes.eps}
  \caption{Search Modes}
  \label{fgr:differentSearchModes}
\end{figure}

@@

%
\begin{acknowledgement}

@@
